---
title: "Data Carpentry Capstone"
author: "Naupaka Zimmerman"
date: "January 12, 2016"
output: html_document
---

```{r setup, include=FALSE}
# Using this, we can set a chunk option for all chunks.
# In this case we don't want to show any code, so this sets 
# echo = FALSE for all chunks unless we set it manually to TRUE
knitr::opts_chunk$set(echo = FALSE)
```

## Task 1: Load the data from a SQLite database

```{r load-packages-and-data, message = FALSE}
# message = FALSE above turns off unwanted output messages about loaded packages

# If needed:
# install.packages("dplyr", dependencies = TRUE)

# The knitr package includes the function kable(), which allows us to format
# tables nicely in our output. The package tidyr includes functions to reformat
# tables from long to wide (see Data Wrangling cheatsheet for details).

# If needed:
# install.packages("knitr", dependencies = TRUE)
# install.packages("tidyr", dependencies = TRUE)

# load packages
library("dplyr")
library("knitr")
library("tidyr")

# 1. Write an Rmd script to load the Portal data from your sqlite database.

################################
# Get data frame from database #
################################

# Set type and location of database
# This database is also available on figshare at
# https://dx.doi.org/10.6084/m9.figshare.1314459
portal_db <- src_sqlite("data/raw/UWM_DC_portal.sqlite")

# Configure SQL query to pull fully joined data from database
to_analyze <- tbl(portal_db, 
				  sql("SELECT sur.*, sp.genus, sp.species, sp.taxa, pl.plot_type
						FROM surveys AS sur
							 JOIN plots AS pl ON sur.plot_id = pl.plot_id
							 JOIN species AS sp ON sp.species_id = sur.species_id
						ORDER BY sur.record_id ASC"))

# Retrieve all data resulting from query into R data frame
collected_data <- collect(to_analyze)
```

```{r display-data-head-as-table}
# Display a small subset fo the data so we know it worked
# and to get a sense for what the data look like
kable(head(collected_data)[1:10], caption = "The first 10 columns of the first few rows of our dataset.")
```

## Task 2: Box plots of weight

```{r calc-means-by-sex}
# create summary data of means by gender for M and F animals
# Subset in case genders other than "M" or "F" are present
mean.weight.by.sex <- collected_data %>% 
    filter(sex == "M" | sex == "F") %>% 
    group_by(sex) %>% 
    summarise(mean = mean(weight, na.rm = TRUE))

# Calculate mean weight across all males, round the result to 2 decimal places
male.mean.weight <- mean.weight.by.sex %>% 
    filter(sex == "M") %>% 
    select(mean)
male.mean.weight <- round(male.mean.weight, 3)

# Calculate mean weight across all females, round the result to 2 decimal places
female.mean.weight <- mean.weight.by.sex %>% 
    filter(sex == "F") %>% 
    select(mean)
female.mean.weight <- round(female.mean.weight, 2)
```

```{r means-t-test}
# Calculate difference between the means with a t-test
# If this is significant, then say so in the markdown text below
# If not, then change to state that it is not significant
# Filter in case genders other than "M" or "F" are present
sex_filtered_data <- collected_data %>%
    filter(sex == "M" | sex == "F") %>%
    select(weight, sex)

weight.ttest <- t.test(weight ~ sex, data = sex_filtered_data, na.rm = TRUE)
```

As you can see in the following figure, the average weight in grams for male species measured was `r male.mean.weight` and for females it was `r female.mean.weight`. This `r if (weight.ttest$p.value < 0.05){"is"} else{"is not"}` a statistically significant difference (p = `r round(weight.ttest$p.value, 2)`).

```{r boxplot-by-sex}
boxplot(weight ~ sex, data = sex_filtered_data, xlab = "Sex", ylab = "Weight in grams")
```

## Task 3: Histogram of all hindfoot lengths

```{r histogram-hindfoot-lengths}
hist(collected_data$hindfoot_length,
	 main = "Histogram of highfoot lengths of all species combined",
	 xlab = "Highfoot length in mm")
```

## Task 4: Scatterplots of weight by hindfoot length

Choose three species. Make three scatterplot of weight (on x) vs hindfoot length for each, with different color points for each species.

### Three selected species

```{r choose-species}
# Choose three species. The three species you set here
# will determine which names are printed out and which 
# species are used for the scatterplots below
chosen.species <- c("DM", "DO", "DS")

# These are the colors for the three scatterplots
chosen.colors <- c("blue", "green", "purple")

# This is to get a small table with the full Genus and species
# for our chosen species_id codes
species.names <- collected_data %>% 
    filter(species_id %in% chosen.species) %>%
    group_by(genus, species) %>% 
    summarise()
```

* *`r paste(species.names$genus[1], species.names$species[1])`*
* *`r paste(species.names$genus[2], species.names$species[2])`*
* *`r paste(species.names$genus[3], species.names$species[3])`*

```{r three-scatterplots-of-weight-by-hindfoot-length}
# Subset and then plot the first species of interest
scatter1 <- collected_data %>%
	filter(species_id == chosen.species[1]) %>%
	select(weight, hindfoot_length)

plot(x = scatter1$weight,
	 y = scatter1$hindfoot_length,
	 xlab = "Weight in grams",
	 ylab = "Hindfoot length in mm",
	 col = chosen.colors[1],
	 main = paste("Weight vs Hindfoot length for species", chosen.species[1]))

# Subset and then plot the second species of interest
scatter2 <- collected_data %>%
	filter(species_id == chosen.species[2]) %>%
	select(weight, hindfoot_length)

plot(x = scatter2$weight,
	 y = scatter2$hindfoot_length,
	 xlab = "Weight in grams",
	 ylab = "Hindfoot length in mm",
	 col = chosen.colors[2],
	 main = paste("Weight vs Hindfoot length for species", chosen.species[2]))

# Subset and then plot the third species of interest
scatter3 <- collected_data %>%
	filter(species_id == chosen.species[3]) %>%
	select(weight, hindfoot_length)

plot(x = scatter3$weight,
	 y = scatter3$hindfoot_length,
	 xlab = "Weight in grams",
	 ylab = "Hindfoot length in mm",
	 col = chosen.colors[3],
	 main = paste("Weight vs Hindfoot length for species", chosen.species[3]))

```


## Task 5: Counts for Species "DM"
For the species code DM, use `dplyr` to calculate the total number of counts per year for each plot type.

```{r number-of-DM-counts-by-year}
# Count up or tally() the number of individuals per plot type
# per year
DM_counts_by_year_and_plot_type <- collected_data %>% 
    filter(species_id == "DM") %>% 
    group_by(year, plot_type) %>% 
    tally()

# We again use kable() to print out a nicely formatted table.
# We also use the spread() function to make the table easier to read, 
# with plto types across the top, and years down the side
kable(spread(DM_counts_by_year_and_plot_type, plot_type, n))

```

## Task 6: Plot counts over time in "Rodent Exclosure" plots
Using the data frame you just created, plot the counts (of the species "DM") over time as line graphs for the plot type “Rodent Exclosure”. Describe what you see.

```{r counts-of-DM-over-time-rodent-exclosure}
# Generate summary table of counts for species "DM" within "Rodent Exclosure"
# plots over time
DM_Rodent_Exclosure_by_year <- collected_data %>% 
    filter(species_id == "DM" & plot_type == "Rodent Exclosure") %>% 
    group_by(year) %>% 
    tally()

# Plot that data with a line graph
plot(x = DM_Rodent_Exclosure_by_year$year, 
	 y = DM_Rodent_Exclosure_by_year$n,
	 type = "l",
	 xlab = "Year",
	 ylab = expression(paste("Count of ", italic("Dipodomys merriami"))))

```

It appears that the Rodent exclosure was effective at reducing counts of *Dipodomys merriami*, particularly after 1990.


